Prelim Outline

### 2019-06-12 draft.tex
need to just finish state of the art and finite size/approximate methods. done!

We propose using a density matrix downfolding (DMD) procedure\textbf{ref} to systematically develop low-energy effective theories of the 1-band nearest-neighbor Hubbard model on a square lattice as a stepping stone towards building effective theories for strongly correlated materials. The Hubbard model is a simple model with nearest-neighbor hopping and a local on-site Coulomb repulsion, parameterized by a single ratio U/t: 
\begin{equation}
H_\text{hub} = -t \sum_{\langle i,j \rangle,\sigma}( c_{i^\dagger,\sigma} c_{j,\sigma} + h.c.) + U \sum_i n_{i\uparrow} n_{i\downarrow}
\label{hubbard}.
\end{equation}
The solutions in 1-d are well understood via the Bethe ansatz \textbf{ref}, however the 2-d square lattice Hubbard model has no known exact solution.  In the very strong coupling limit $U/t \rightarrow \infty$ an emergent low-energy spin-space exists with an effective nearest-neighbor AFM Heisenberg Hamiltonian \textbf{ref}. In the region $U = 0$ a trivial non-interacting theory holds. Between these two limits, in particular the intermediate coupling region $1 < U < 7$, no well established effective theory exists. In this region there is a metal-insulator transition at $U_c/t \sim 4$ \textbf{ref}, and the pseudogap opens up at finite temperature around $U/t \sim 7$, leading to the development of a Mott gap for large $U/t > 11$ \textbf{ref}. In addition to the half-filling phase diagram, this model exhibits a plethora of interesting ground state properties when doped including stripe phases and d-wave pairing \textbf{ref}. The relative simplicity of the Hubbard model and the lack of a well-established low-energy theory in a broad intermediate coupling region lends the model as a useful playground for our new downfolding procedure in preparation for tackling realistic strongly correlated electron systems.

@@

\section{Proposed work}
We want to develop low-energy effective theories for the
half-filled Hubbard model over the space U: 0 - 12, 1/k$_B$T: 0 - 3 using the DMD method, particularly focusing on the medium coupling region of U: 2 - 6 and 1/k$_B$T $\sim$ 1 where perturbative effective theories break down and a metal-insulator transition exists.
Similar to the case of the \textit{ab-initio} downfolding, we will need to construct low-energy spaces, sample the constructed spaces, build effective theories and respective bases, and fit our models.
Unlike \textit{ab-initio} downfolding, the Hubbard model downfolding has significantly fewer degrees of freedom, and therefore the DMD procedure can be done in a much more controlled fashion.
We will begin from a starting point of highest control, namely for system sizes small enough where we can exactly diagonalize the Hubbard model.
In the situation where we have full access to the eigenstates and spectra of a system, selecting $\mathcal{LE}$ can be done exactly as the span of some chosen $N$ lowest energy eigenstates. 
We can further guarantee that our sample set fully encapsulates all low-energy degrees of freedom by simply sampling states parameterized as $|\Psi\rangle = \sum_{j=1}^N c_j |\Phi_j\rangle$ where $|\Phi_j\rangle$ are eigenstates of the Hubbard Hamiltonian.
This can be seen in contrast to the CuO fitting where $\mathcal{LE}$ was defined implicitly through a number occupation descriptor and the sampling of $\mathcal{LE}$ was incomplete leading to intruder states in our initial effective theories.
Basis construction for the Hubbard Hamiltonian is also explicitly clear as any possible single particle basis we can use is just an orthogonal transformation of the basis the Hubbard model is defined on, and is guaranteed to be a complete single particle basis.
Again, this can be compared to the case of an \textit{ab-initio} system like the CuO molecule where bases are only approximately complete due to the huge number of degrees of freedom in the system.
In order to include temperature into the mix we can work with the full model Hilbert space such that $\mathcal{LE} = \mathcal{H}$, but generate samples within our space using a Monte Carlo sampling method according to a Boltzmann probability distribution with variable parameter $\beta = 1/k_BT$.
This way, our sample space will still be able to probe every excitation within a certain temperature range, the samples will be biased towards lower energy states according to the Boltzmann weight, and hence the fit model will also be biased towards describing more accurately the energy functional on the states with highest weight.
The Monte Carlo sampling may need to be altered further in order to deal with entropic effects when the density of states is very low in an energy range of interest and relatively high elsewhere.

Show in \ref{fig:Hubbard} is ... \textbf{SHOW OUR PRELIMINARY RESULTS}

In order to make any statements about effective theories for the Hubbard model we will have to work with larger system sizes and approximate methods like lattice quantum Monte Carlo to generate low-energy states for the DMD procedure. \textbf{TALK ABOUT FINITE SIZE SCALING AND APPROXIMATE METHODS (liken to FN-DMC for ab initio)}

A farther reaching goal would be to develop effective theories in the U-T-$\delta$ (doping) space for the Hubbard model, where more complex behavior similar to the cuprates superconductors have been found.

>>>>

We propose using a density matrix downfolding (DMD) procedure\textbf{ref} to systematically develop low-energy effective theories of the 1-band nearest-neighbor Hubbard model on a square lattice as a stepping stone towards building effective theories for strongly correlated materials. The Hubbard model is a simple model with nearest-neighbor hopping and a local on-site Coulomb repulsion, parameterized by a single ratio U/t: 
\begin{equation}
H_\text{hub} = -t \sum_{\langle i,j \rangle,\sigma}( c_{i^\dagger,\sigma} c_{j,\sigma} + h.c.) + U \sum_i n_{i\uparrow} n_{i\downarrow}
\label{hubbard}.
\end{equation}
The solutions in 1-d are well understood via the Bethe ansatz \textbf{ref}, however the 2-d square lattice Hubbard model has no known exact solution.  In the very strong coupling limit $U/t \rightarrow \infty$ an emergent low-energy spin-space exists with an effective nearest-neighbor AFM Heisenberg Hamiltonian \textbf{ref}. In the region $U = 0$ a trivial non-interacting theory holds. Between these two limits, in particular the intermediate coupling region $2 < U < 8$, no well established effective theory exists. In this region there is a metal-insulator transition at $U_c/t \sim 4$ \textbf{ref}, and the pseudogap opens up at finite temperature around $U/t \sim 7$, leading to the development of a Mott gap for large $U/t > 11$ \textbf{ref}. In addition to the half-filling phase diagram, this model exhibits a plethora of interesting ground state properties when doped including stripe phases and d-wave pairing \textbf{ref}. The relative simplicity of the Hubbard model and the lack of a well-established low-energy theory in a broad intermediate coupling region lends the model as a useful playground for our new downfolding procedure in preparation for tackling realistic strongly correlated electron systems.

@@

\section{Proposed work}
Our goal is to develop low-energy effective theories for the
Hubbard model over the space U-T-$\delta$ using the DMD method, particularly focusing on the medium coupling regions and moderate dopings where perturbative effective theories break down.
Similar to the case of the \textit{ab-initio} downfolding, we will need to construct low-energy spaces, sample the constructed spaces, build effective theories and respective bases, and fit our models.
Unlike \textit{ab-initio} downfolding, the Hubbard model downfolding has significantly fewer degrees of freedom, and therefore the DMD procedure can be done in a much more controlled fashion.
We will begin from a starting point of highest control, namely for system sizes small enough where we can exactly diagonalize the Hubbard model.
In the situation where we have full access to the eigenstates and spectra of a system, selecting $\mathcal{LE}$ can be done exactly as the span of some chosen $N$ lowest energy eigenstates. 
We can further guarantee that our sample set fully encapsulates all low-energy degrees of freedom by simply sampling states parameterized as $|\Psi\rangle = \sum_{j=1}^N c_j |\Phi_j\rangle$ where $|\Phi_j\rangle$ are eigenstates of the Hubbard Hamiltonian.
This can be seen in contrast to the CuO fitting where $\mathcal{LE}$ was defined implicitly through a number occupation descriptor and the sampling of $\mathcal{LE}$ was incomplete leading to intruder states in our initial effective theories.
Basis construction for the Hubbard Hamiltonian is also explicitly clear as any possible single particle basis we can use is just an orthogonal transformation of the basis the Hubbard model is defined on, and is guaranteed to be a complete single particle basis.
Again, this can be compared to the case of an \textit{ab-initio} system like the CuO molecule where bases are only approximately complete due to the huge number of degrees of freedom in the system.
In order to include temperature into the mix we can work with the full model Hilbert space such that $\mathcal{LE} = \mathcal{H}$, but generate samples within our space using a Monte Carlo sampling method according to a Boltzmann probability distribution with variable parameter $\beta = 1/k_BT$.
This way, our sample space will still be able to probe every excitation within a certain temperature range, the samples will be biased towards lower energy states according to the Boltzmann weight, and hence the fit model will also be biased towards describing more accurately the energy functional on the states with highest weight.
The Monte Carlo sampling may need to be altered further in order to deal with entropic effects when the density of states is very low in an energy range of interest and relatively high elsewhere.

In order to make any statements about effective theories for the Hubbard model we will have to work with larger system sizes and approximate methods like lattice quantum Monte Carlo to generate low-energy states for the DMD procedure.

We conclude by showing some very preliminary results on model fitting for the half-filled, 2x2 square Hubbard model for U: 0 - 12 which put our approach to model fitting with exact diagonalization at T$\geq 0$ into practice.
We start off by exactly diagonalizing each model, define $\mathcal{LE} = \mathcal{H}$, and treat as our full sample space as just the eigenstates of the model.
We then fit three single parameter models to the eigenstates: A) 
A "t" only model $t \sum_{\langle i, j \rangle, \sigma} c_{i\sigma}^\dagger c_{j\sigma} + h.c.$, B) A "U" only model $U \sum_{i} n_{i\uparrow} n_{i\downarrow}$, and C) A "J" only model $J \sum_{\langle i,j \rangle} \vec{S}_i \cdot \vec{S}_j$ using a weighted linear regression with weights $e^{-\beta E}$.
The basis for all of these operators are just the bare Hubbard Hamiltonian basis.
We have restricted ourselves to eigenstates only as we have not developed a sufficient low-energy sampling scheme yet which deals with the entropy issue discussed earlier.
The weighted linear regression approximates the effect of generating samples using a Markov Chain Monte Carlo method with a Boltzmann distribution.
For small U the t-only model is an accurate effective theory as the temperature increases, which is to be expected in this perturbative limit.
If one were to look at large U and high temperature a U-only model is quite good since the dominant and relevant energy scale at higher temperatures is related to double occupation.
At large U and low temperature we see that the expected Heisenberg model accurately describes the low-energy excitations. 
In this region a t-only model also works quite accurately since for large U the hopping descriptor $\sum_{\langle i, j \rangle, \sigma} c_{i\sigma}^\dagger c_{j\sigma} + h.c.$ is correlated to the exchange operator $J \sum_{\langle i,j \rangle} \vec{S}_i \cdot \vec{S}_j$ for the lowest energy excitations in the system.
It should be noted, however, that this t-only model would have to be explicitly written to only act on states with no double occupancy, an assumption inherent in the Heisenberg model. 
Finally, in the region beginning at 4 $<$ U $<$ 8 and moderate $\beta$: 0.1 - 0.4 we see that none of the candidate single parameter models can accurately describe the low-energy excitations of the system.
This poorly explained region indicates the breakdown of perturbative expansions of the Hubbard model about the strong- or weak-coupling limits.
In order to make more conclusive statements we would have to actually sample the Hilbert space to construct a more robust estimate of the best models for different $U, \beta$.
We would be particularly interested in whether an effective theory for this region exists and how this region and corresponding effective theory change as we increase the system size.

\begin{figure}[H]
\centering
\includegraphics[width=0.7\linewidth]{../hubbard/eff_eig_prelim.pdf}
\caption{Results of weighted single parameter model fits to the eigenstates of a Hubbard model for various U and temperature. Darker hues indicate a higher R$^2$ score.}
\label{fig:Hubbard}
\end{figure}
